---
title: "Introduction to daa"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction to daa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The **daa** package provides methods for differential abundance analysis (DAA)
of microbiome data. It works with `TreeSummarizedExperiment` and 
`SummarizedExperiment` objects from the miaverse ecosystem.

## Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("daa")
```

# Available Methods

## Wilcoxon rank-sum test

The `getWilcoxonTest()` function performs nonparametric Wilcoxon rank-sum tests
for each feature, comparing two groups.

```{r wilcoxon, message = FALSE}
library(daa)
library(mia)

# Load example data
data(peerj13075, package = "mia")
tse <- peerj13075

# Agglomerate to phylum level
tse <- agglomerateByRank(tse, "phylum")

# Run Wilcoxon test
res <- getWilcoxonTest(tse, counts ~ Geographical_location)
head(res)
```

## t-test

The `getTTest()` function performs t-tests (Welch's by default).

```{r ttest}
res <- getTTest(tse, counts ~ Geographical_location)
head(res)
```

## ANOVA and Kruskal-Wallis

For comparing more than two groups:

```{r anova}
res <- getAnova(tse, counts ~ Geographical_location)
head(res)
```

```{r kruskal}
res <- getKruskal(tse, counts ~ Geographical_location)
head(res)
```

## Testing metadata variables

You can also test variables from `colData`. For example, testing
alpha diversity indices:

```{r metadata}
# Add alpha diversity to colData
tse <- mia::addAlpha(tse, index = "shannon")

# Test diversity differences between groups
res <- getWilcoxonTest(tse, shannon ~ Geographical_location)
head(res)
```

## Wide-format methods

For model-based approaches, use formula with the assay name on the LHS:

```{r lm, eval = FALSE}
# Linear model per feature
tse <- transformAssay(tse, method = "rclr")
res <- getLm(tse, rclr ~ Geographical_location)

# Limma
fit <- getLimma(tse, counts ~ Geographical_location)
```

# Session Info

```{r session}
sessionInfo()
```
