---
title: "Introduction to daa"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction to daa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The **daa** package provides methods for differential abundance analysis (DAA)
of microbiome data. It works with `TreeSummarizedExperiment` and 
`SummarizedExperiment` objects from the miaverse ecosystem.

## Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("daa")
```

# Available Methods

## Wilcoxon rank-sum test

The `getWilcoxon()` function performs nonparametric Wilcoxon rank-sum tests
for each feature, comparing two groups.

```{r wilcoxon, message = FALSE}
library(daa)
library(mia)

# Load example data
data(GlobalPatterns)
tse <- GlobalPatterns

# Subset to 2 groups and first 50 features for speed
tse <- tse[1:50, tse$SampleType %in% c("Feces", "Skin")]

# Run Wilcoxon test
res <- getWilcoxon(tse, assay.type = "counts", formula = x ~ SampleType)
head(res)
```

Use `addWilcoxon()` to store results in `rowData()`:

```{r add_wilcoxon}
tse <- addWilcoxon(tse, assay.type = "counts", formula = x ~ SampleType)
head(rowData(tse)[, "wilcoxon_pvalue", drop = FALSE])
```

## t-test

The `getTtest()` function performs t-tests (Welch's by default).

```{r ttest}
res <- getTtest(tse, assay.type = "counts", formula = x ~ SampleType)
head(res)
```

# Session Info

```{r session}
sessionInfo()
```
